---
title: "04. Code maps and dual-mode development"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{04. Code maps and dual-mode development}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

FG Studio must remain understandable and maintainable by humans **without an AI
assistant**, while agents must be able to find the same source efficiently.
Generated diagrams supplement the reviewed call explanations in articles 01–03;
they do not replace those explanations or establish completeness.

## Application dependencies with flow

This bounded diagram is generated from the current package. It describes static
dependencies, not the chronological execution of every reactive callback.

```{r app-dependencies, echo=TRUE, message=FALSE, warning=FALSE}
diagram <- flow::flow_view_deps(fgstudio::fgstudio_app, max_depth = 2,
  show_imports = "packages")
diagram$widget
```

## Inside one function

`valid_map_bounds` validates saved map-view coordinates. This small example shows
how a developer can inspect decision logic without rendering an entire package.
It does not perform scientific spatial containment; that belongs in fluvgeo.

```{r bounds-logic, echo=TRUE, message=FALSE, warning=FALSE}
flow::flow_view(fgstudio:::valid_map_bounds)$widget
```

For larger functions, start with their explanation and source, then request a
bounded diagram. A giant graph is not a substitute for a comprehensible design.
We embed flow's native HTML widget: flow 0.2.0's default knitr printer tries to
create a PNG through webshot, which failed on this workstation. The native widget
avoids that screenshot dependency without replacing the graph generator.

## Where an automatic graph needs interpretation

| Relationship | How to interpret it |
|---|---|
| A regular named function call | Candidate static dependency; inspect its source |
| `store$split_reach(...)` | A closure returned by `local_study_store`, not a package-level function |
| `revise(..., writer=fluvgeo::split_study_reach)` | A function supplied as an argument and invoked indirectly |
| `on_saved(...)` | A callback supplied by the parent module that updates current study state |
| Shiny observers | Registered now, invoked later; call order is not event order |
| `callr::r_bg(drainage_request, ...)` | A worker-process boundary, not an ordinary synchronous call |

`pkgnet` provides within-package function-network data suitable for export and
querying. Its documented limitations include list-held functions, some method
calls and nonstandard evaluation. In particular, absence of `store$split_reach`
from a network does not mean the app never calls the backend split function.
See the [pkgnet reporter contract](https://uptake.github.io/pkgnet/reference/FunctionReporter.html).

## Updating both documentation modes

When adding a capability, extend the appropriate article or add the next numbered
article. Explain entry events, calls, indirect dispatch, state, persistence,
failure boundaries and tests. Keep the short agent route in step with it. A human
should be able to trace UI → adapter → backend → saved output without recovering
the design from a chat transcript.

Use the repository's `dev/workflows/developer-documentation.md` for rebuilding,
checking and navigating the machine-readable index. Generated graph evidence must
identify its source snapshot and limitations; code and tests remain authoritative.
No automatic call graph is claimed to enumerate all possible runtime behavior.
