---
title: "01. Code maps and dual-mode development"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{01. Code maps and dual-mode development}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

FG Studio must remain understandable and maintainable by humans **without an AI
assistant**, while agents must be able to find the same source efficiently.
The [project overview](fgstudio.html) explains the analyst purpose this work serves.
Generated diagrams supplement the reviewed call explanations across the developer series;
they do not replace those explanations or establish completeness.

## Choose a route for the question

Context routing means finding the smallest useful set of current explanations,
contracts, source and tests for a task. Reading every document is not the goal.
Human and agent routes should converge on the same evidence, while remaining
usable independently. An analyst does not need this layer to operate the app.

| Question | Human entry | Repository/agent entry | Verify against |
| --- | --- | --- | --- |
| What belongs in this app or the backend? | [Project overview](fgstudio.html), then article 02 | `dev/architecture/design.md`; FG-architecture for cross-repository work | Member repository contracts, current source and tests |
| Why does a save or reload behave this way? | [Lifecycle](dev-02-application-lifecycle.html), then the relevant feature article | `dev/architecture/agent-routes.md` | Module, adapter, backend writer/reader and regression test |
| How does a Reach edit cross the backend boundary? | [Reach editing](dev-04-reach-editing.html) | The Reach row in the route table | Callback, injected writer, persisted output and backend tests |
| What is the state of CRS or DEM development? | Articles [06](dev-06-stream-dem-files.html), [07](dev-07-dem-inspection.html), [08](dev-08-analysis-crs.html), [09](dev-09-vertical-reference.html) | Matching feature and current project plan | Actual UI entry points and tested capability limits |
| Which function calls this one? | A bounded diagram or source search | Query the fresh code map for the named symbol | Source; reviewed bridges for indirect calls |
| Did this documentation improve maintenance efficiency? | The pilot and comparison method below | `dev/governance/navigation-pilot-2026-09-19.md` | Correct answers, comparable measurements and explicit limitations |

Paths in this article are relative to the fgstudio checkout. The
[repository route table](https://github.com/FluvialGeomorph/fgstudio/blob/main/dev/architecture/agent-routes.md)
provides precise source/test entry points. Use the local copy during development;
published links may lag local changes. Code and tests establish actual behavior,
accepted decisions govern design, and schemas define documented contracts.

## What the context-routing experiment has established

The [first pilot](https://github.com/FluvialGeomorph/fgstudio/blob/main/dev/governance/navigation-pilot-2026-09-19.md)
compared two fresh-context agents on the same three read-only questions: Reach
split persistence, Stream clipping ownership, and Reach merge blockers. One used
source-only navigation; the other could use routes, articles and generated aids.
An independent source review checked the decisive answers.

Both answered the core questions correctly. The documentation-assisted run read
more test evidence and used the route table plus two articles; it did not use the
generated graph. Self-reported shell calls were 7 versus 6, and files read were
12 versus 16. Those counts did not measure equivalent reading depth. Complete
elapsed times and token consumption were not recorded comparably.

The finding supports retaining concise routes and explanations. It **does not
establish faster work, lower context consumption, a measured human benefit or
an independent benefit from the graph**. Human maintainability remains an explicit
requirement to evaluate, not an outcome proven by an agent-only pilot. Runtime
tests remain separate evidence of application behavior.

## How to continue evaluating the method

Use real maintenance questions when a navigation problem warrants a comparison;
do not rerun a benchmark for every wording change. Hold the source snapshot,
question, expected answer, allowed tools and verification depth constant. Compare
source navigation with the proposed aids, using fresh context and independent
review. Include human maintenance tasks when assessing the human requirement.

Record correctness and decisive citations first, then complete elapsed time,
actual context/tokens where available, files inspected, irrelevant detours and
documentation-maintenance effort. Mark unavailable measures as unknown. A lower
call count alone is not a saving. Separate the effects of route tables, prose and
graphs where possible; record failures and negative results as well as successes.

Retain small aids that repeatedly help a real task. Simplify unclear code before
building a larger graph around it. Promote methods to shared **reproducibleai**
guidance only after repeated evidence, through that project's own review. This
app's experiment does not impose a workspace-wide documentation framework.

## Machine-readable navigation without a second specification

The local site includes article Markdown and `llms.txt`; these are text entry
points to the same documentation, not agent instructions or separate scientific
authority. The repository's `dev/architecture/call-network.json` records static
edges, reviewed indirect bridges and source hashes. Check freshness, then query
a specific symbol rather than loading the entire graph into every task:

```text
Rscript --vanilla dev/scripts/query-code-map.R --check
Rscript --vanilla dev/scripts/query-code-map.R fluvgeo::split_study_reach
```

Resolve R through the workstation instructions when it is not on PATH. A missing
edge does not prove a call is absent; source hashes establish freshness, not
semantic completeness. No AI service is part of the user workflow. Development
agents are optional collaborators, and human review retains scientific authority.

## Where implementation and verification belong

| Location | Responsibility |
| --- | --- |
| `R/` | Installed application functions, internal helpers and runtime input validation; shared scientific methods belong in fluvgeo |
| `tests/testthat/` | Repeatable package behavior and regression tests, with controlled fixtures and temporary outputs |
| `dev/scripts/` | Build/run tooling, documentation checks, explicit live-service or workstation-data qualification and cross-repository orchestration |
| `dev/check-output/` | Ignored local evidence and temporary experiment output; not the authoritative regression suite |

The package uses the standard `tests/testthat.R` entry point. The `dev/` directory
is excluded by `.Rbuildignore`, so its standalone assertions do not run as part of
`R CMD check`. That is intentional for checks requiring public services, real
local studies or documentation-build artifacts. It must not become a way to leave
ordinary package regressions outside automated verification. Live tests can also
be organized as explicitly gated integration tests; their location is a project
choice, not an R prohibition.

JavaScript runtime contracts now run through `test-javascript.R`, with bounded
Node fixtures under `tests/testthat/javascript/`. R supplies the actual widget
callbacks or installed asset, so they also work when checking a source package.
Node.js 18 or newer must be on PATH for the maintained development check runners;
ordinary package tests report explicit skips if Node is unavailable. Automation
can enforce complete coverage with `FGSTUDIO_REQUIRE_NODE=true`. These controlled
DOM/map fixtures complement browser acceptance, which still requires a browser.

Use standard R development commands first, following
`dev/workflows/r-package-development.md`. Existing scripts such as `check-tests.R`
still provide extra orchestration and remain under review; their presence does
not establish that they are necessary. Before adding a wrapper, check the standard
tool's options and identify the specific gap. Public-service qualification remains
opt-in. Promote stable regression cases from reproductions into package tests.

See `dev/scripts/README.md` for the current classification and
`dev/workflows/r-package-development.md` for placement rules. This follows the
[R Packages testing guidance](https://r-pkgs.org/testing-design.html): production
functions and test expectations have different execution lifecycles. Tests should
not run when an analyst loads the application.

## Application dependencies with flow

This bounded diagram is generated from the current package. It describes static
dependencies, not the chronological execution of every reactive callback.

```{r app-dependencies, echo=TRUE, message=FALSE, warning=FALSE}
diagram <- flow::flow_view_deps(fgstudio::fgstudio_app, max_depth = 2,
  show_imports = "packages")
diagram$widget
```

## Inside one function

`valid_map_bounds` validates saved map-view coordinates. This small example shows
how a developer can inspect decision logic without rendering an entire package.
It does not perform scientific spatial containment; that belongs in fluvgeo.

```{r bounds-logic, echo=TRUE, message=FALSE, warning=FALSE}
flow::flow_view(fgstudio:::valid_map_bounds)$widget
```

For larger functions, start with their explanation and source, then request a
bounded diagram. A giant graph is not a substitute for a comprehensible design.
We embed flow's native HTML widget: flow 0.2.0's default knitr printer tries to
create a PNG through webshot, which failed on this workstation. The native widget
avoids that screenshot dependency without replacing the graph generator.

## Where an automatic graph needs interpretation

| Relationship | How to interpret it |
|---|---|
| A regular named function call | Candidate static dependency; inspect its source |
| `store$split_reach(...)` | A closure returned by `local_study_store`, not a package-level function |
| `revise(..., writer=fluvgeo::split_study_reach)` | A function supplied as an argument and invoked indirectly |
| `on_saved(...)` | A callback supplied by the parent module that updates current study state |
| Shiny observers | Registered now, invoked later; call order is not event order |
| `callr::r_bg(drainage_request, ...)` | A worker-process boundary, not an ordinary synchronous call |

`pkgnet` provides within-package function-network data suitable for export and
querying. Its documented limitations include list-held functions, some method
calls and nonstandard evaluation. In particular, absence of `store$split_reach`
from a network does not mean the app never calls the backend split function.
See the [pkgnet reporter contract](https://uptake.github.io/pkgnet/reference/FunctionReporter.html).

## Updating both documentation modes

When adding a capability, extend the appropriate article or add the next numbered
article. Explain entry events, calls, indirect dispatch, state, persistence,
failure boundaries and tests. Keep the short agent route in step with it. A human
should be able to trace UI → adapter → backend → saved output without recovering
the design from a chat transcript.

Use the repository's `dev/workflows/developer-documentation.md` for rebuilding,
checking and navigating the machine-readable index. Generated graph evidence must
identify its source snapshot and limitations; code and tests remain authoritative.
No automatic call graph is claimed to enumerate all possible runtime behavior.
